---
title: "Evaluating Cardiovascular and Metabolic Predictors of 1 - Year Dementia Risk"
author: "Rithika Saravanan"
date: "May 2, 2026"
format: pdf
execute: 
  warning: false
  message: false
  echo: false
editor: visual
---

```{r}
#import datasets & libraries
library(tidyverse)
library(tidymodels)
library(knitr)
library(janitor)

patient_risk_profiles <- read.csv("data/patient_risk_profiles.csv")

#clean all variable names in DF to create unique identifiers
df_clean <- patient_risk_profiles %>%
  clean_names()

#Collapse all age columns to only 1 with midpoint formula
df_final <- df_clean %>%
  pivot_longer( #long format
    cols = starts_with("age_group"), names_to = "age_range", 
    values_to = "is_present"
  ) %>%
  #filter rows with patient in age group
  filter(is_present == 1) %>%
  #get high and low end of range ==> midpoint
  mutate(
    age_low = as.numeric(str_extract(age_range, "\\d+")),
    age_high = as.numeric(str_extract(age_range, "\\d+$")),
    age_numeric = (age_low + age_high) / 2
  ) %>%
  #remove all old columns
  select(-age_range, -is_present, -age_low, -age_high)

#collapse sex columns ==> sex (0 = Male, 1 = Female)
df_final <- df_final %>%
  mutate(sex = case_when(
    sex_female == 1 ~ 1,
    sex_male == 1 ~ 0
  )) %>%
  #remove old columns
  select(-sex_female, -sex_male)

#combine stroke columns
df_final <- df_final %>%
  mutate(any_stroke = ifelse(
    hemorrhagic_stroke_in_an_inpatient_setting_in_prior_year == 1 | 
    non_hemorrhagic_stroke_in_an_inpatient_setting_in_prior_year == 1, 
    1, 0)) %>%
  select(-hemorrhagic_stroke_in_an_inpatient_setting_in_prior_year,
         -non_hemorrhagic_stroke_in_an_inpatient_setting_in_prior_year)

#str(df_final)

#shorten column names & filter df to only useful predictors
df_an <- df_final %>%
  rename(
    dementia_risk = predicted_risk_of_dementia,
    hypertension = hypertension_in_prior_year,
    hyperlipidemia = hyperlipidemia_in_prior_year,
    smoking = smoking_in_prior_year,
    t2_diabetes = type_2_diabetes_mellitus_dm_with_no_type_1_or_secondary_dm_in_prior_year,
  ) %>%
  #filter only useful col
  select(person_id, age_numeric, sex, dementia_risk, 
         hypertension, hyperlipidemia, any_stroke, smoking, t2_diabetes)

# Convert binary indicators to factors with labels
df_an <- df_an %>%
  mutate(
    sex = factor(sex, levels = c(0, 1), labels = c("Male", "Female")),
    hypertension = factor(hypertension, levels = c(0, 1), labels = c("No", "Yes")),
    hyperlipidemia = factor(hyperlipidemia, levels = c(0, 1), labels = c("No", "Yes")),
    any_stroke = factor(any_stroke, levels = c(0, 1), labels = c("No", "Yes")),
    smoking = factor(smoking, levels = c(0, 1), labels = c("No", "Yes")),
    t2_diabetes = factor(t2_diabetes, levels = c(0, 1), labels = c("No", "Yes"))
  )

#Model 1: Demographics only
m1 <- lm(dementia_risk ~ age_numeric + sex, data = df_an)

# Model 2: Adding Vascular Indicators (The Core Hypothesis)
m2 <- lm(dementia_risk ~ age_numeric + sex + hypertension + any_stroke + hyperlipidemia, data = df_an)

# Model 3: The Full Predictive Model
m3 <- lm(dementia_risk ~ age_numeric + sex + hypertension + any_stroke + hyperlipidemia + t2_diabetes + smoking, data = df_an)

# Model 4: Vascular Indicators + Interaction
m4 <- lm(dementia_risk ~ sex + + any_stroke + hyperlipidemia + age_numeric * hypertension, data = df_an)

#Model 5: Vascular Indicators + Log-transformed(dementia_risk)
m5<- lm(log(dementia_risk) ~ age_numeric + sex + hypertension + any_stroke + hyperlipidemia, data = df_an)
```

## Introduction

Dementia is a growing public health concern in the 21st century, especially with the aging population. It is imperative that we understand the factors associated with dementia risk to help identify and assist high-risk individuals for early intervention. The modern "Vascular Hypothesis" suggests that cognitive decline and dementia are linked to systematic cardiovascular health. So this study aims to quantify how these vascular health predictors, like hypertension, hyperlipidemia, and history of smoking behavior, influence an individual's predicted 1-year risk of dementia.

The data used in this analysis is from TidyTuesday Patient Risk Profiles dataset (October 2023), originally prepared for the R/Pharma conference and contributed by Jenna Reps. The link to the dataset is: <https://github.com/rfordatascience/tidytuesday/blob/main/data/2023/2023-10-24/readme.md>. It consists of 100 simulated patient profiles generated from real-world healthcare data and logistic regression models. Each observation (row) represents a unique simulated patient (represented by personId), containing their medical history and predicted 1-year risk scores for 14 health outcomes. The observational unit is an individual (simulated) patient. Initial data cleaning and analysis found no missing NA values in the dataset (refer to Appendix Figure I). In addition, to ensure a reproducible and principled analysis, several cleaning steps were performed:

-   **Standardization:** Non-standard raw column names were converted to snake_case using the `janitor` package.

-   **Age Continuous Transformation:** 20 Age intervals (e.g., 70-74) variables were converted to a singular variable using the **midpoint formula** for each patient.

-   **Feature Engineering:** Hemorrhagic and non-hemorrhagic stroke indicators were combined into a single "Any Stroke" variable to strengthen the clinical signal. Similarly, sex columns were combined to a single factor (Male/Female). Key variables were renamed for clarity and a new dataframe was created with the subset of binary indicated converted to labeled factors we focus on.

**Research Question**: What patient characteristics (demographic and medical history factors) are most strongly associated with higher predicted 1-year dementia risk in this simulated cohort?

The response variable of interest is predicted_risk_of_dementia (a probability between 0 and 1). Initial EDA analysis of Dementia Risk among the patients demonstrated a right-skew of the data, so majority of the patients are low-risk (refer to Figure 1 below). Predictors include age (collapsed from age group indicators), sex, and several binary comorbidity/smoking indicators such as hypertension, hyperlipidemia, any stroke (hemorrhagic or non-hemorrhagic), type 2 diabetes, and smoking history in the prior year. Moreover, Figure 2 confirms our sample spans across the average full adult lifecycle, with adult midpoints every 5 years, to provide a good baseline for testing age-dependent dementia risk.

```{r}
#EDA of response variable - dementia_risk
ggplot(df_an, aes(x = dementia_risk)) +
  geom_histogram(bins = 50, color = "white") +
  labs(title = "Figure 1 - Distribution of Predicted 1-Year Dementia Risk",
       x = "Predicted Risk Score (0-1)",
       y = "Count")
```

```{r}
#Univariate EDA of predictor variable - age
ggplot(df_an, aes(x = age_numeric)) +
  geom_histogram(binwidth = 5, color = "white") +
  labs(title = "Figure 2 - Distribution of Patient Ages",
       x = "Ages (Midpoint of 5−year bracket)",
       y = "Count")
```

# Methodology

Given that the response variable (predicted 1-year dementia risk) is a continuous numeric variable and our goal is to quantify the individual and combined contributions multiple clinical vascular variables play, Multiple Linear Regression was selected as the most appropriate approach. MLR allows for the isolation of specific, individual variable effects and the combined interactive effects, while also controlling for essential confounding variables (ex. age and sex). This method of analysis was chosen over others because the primary goal is to analyze the magnitude of dementia risk, rather than a simple binary "at risk/not at risk" categorization like with logistic regression.

A nested modeling approach was used to create the most effective statistical model to predict dementia risk.

-   Model 1 was the baseline demographic model that just used the patient's age and sex to predict dementia risk.

-   Model 2 added the vascular predictors, so used a patient's age, sex, hypertension history, stroke history, and hyperlipidemia history to predict their dementia risk score.

-   Model 3 was the full model that used a patient's age, sex, hypertension history, stroke history, hyperlipidemia history, type 2 diabetes history, and smoking history to predict their dementia risk score.

-   Model 4 added an interaction term between age and hypertension history in addition to a patient's stroke history, hyperlipidemia history, and sex to predict one's dementia risk. This approach lets one see the incremental predictive gain and the complexity increases.

-   Model 5 checks how well a patient's age, sex, hypertension history, stroke history, and hyperlipidemia history predict the log-transformed dementia_risk of a patient, since the main response variable is also skewed towards 0 (refer to Figure 1).

Ultimately, the best model was seen to be model 5 (will be explained why in the Results section). In addition, all assumptions of linear regression were verified for the final chosen model (model 5). Refer to Appendix Figure II for more details on the analysis performed in R:

-   **Linearity & Constant Variance:** The Residuals v. Fitted value plot showed a random scatter of points around 0 and a constant spread of residuals.

-   **Independence**: Since each observation represented a unique simulated patient profile with no mentioning of clustering, we are safe to assume independence of outcomes.

-   **Normality**: A Q-Q plot shows that the residuals follow closely along the reference line.

## Results

Summary statistics of the dataset revealed that around 16% of patients have a history with hypertension, 14% show hyperlipidemia, and 27% have experienced some type of stroke. Moreover, the mean age of the patients was 46.8 years (refer to Appendix Figure III). This gives context for the different models compared in this study.

```{r}
cat("Figure 3:  Model Comparison Statistics")
model_comp <- data.frame(
  "Model" = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
  "R^2" = c(0.292, 0.418, 0.425, 0.421, 0.921),
  "Adjusted R^2" = c(0.277,  0.388, 0.381, 0.384, 0.917),
  "Residual Standard Error (RSE)" = c(0.061, 0.056, 0.057, 0.056, 0.765),
  "F-Statistic" = c(19.99, 13.55, 9.71, 11.26, 219.40))
kable(model_comp)

```

Refer to Figure 3 above to see how the different models compare in predictive power. Among the baseline demographic model (model 1), vascular predictors model (model 2), and full model with smoking and type 2 diabetes history (model 3), model 1 had the lowest adjusted R\^2 value (0.277) and model 2 had the highest R\^2 value (0.388). To test if an interaction term between age and hypertension history would add significant predictive power, model 4 incorporated the vascular predictors and this interaction term. However, the adjusted R\^2 for model 4 was slightly lower than model 2 even though the R\^2 value for model 4 was higher than the R\^2 value for model 2. Thus, adding the added complexity of adding the interaction term between age and hypertension history did not significantly improve the predictive power of the model.

Nevertheless, when checking if model 2 complied with the assumptions for linear regression, it was found that the Residuals vs. Fitted plot violates the linearity and constant variance conditions by "fanning out" and clustering more on the left side than the right (refer to Appendix Figure IV). This is likely due to the main response variable dementia_risk being skewed towards 0 (refer to Figure 1). By log-transforming the response variable in model 5, the variance was stabilized (refer to Appendix Figure II) and achieved an adjusted R-squared value of around 0.917, meaning that the clinical variables in the model explain around 92% of the variance in dementia risk.

```{r}
cat("Figure 4: Model 5 Statistics")
tidy(m5) |>
  kable(digits = 3)
```

The final model is: $log(dementia \ risk) = \beta_0 + \beta_1(Age) + \beta_2(sex) + \beta_3(hypertension) + \beta_4(stroke) + \beta_5(hyperlipidemia) + \epsilon$In the final model (Model 5), a patient's age and whether they have had a stroke appear to be the most statistically significant clinical predictors (p \< 0.05) in predicting one's dementia_risk score. Because the model is log-transformed, the coefficients in Figure 4 represent the multiplicative change in one's dementia risk score. For example, when controlling for all other variables in the model, a history of stroke is associated with a statistically higher predicted dementia risk score for a patient compared to those without a stroke history, on average. Similarly, one's age increasing is associated with a steady increase in the log-risk of dementia, when controlling for all other variables in the model, on average. On the other hand, hypertension, hyperlipidemia, and sex are not statistically significant predictors of dementia risk, when controlling for all other variables in the model. The boxplot of dementia risk score versus Stroke history in Appendix figure V confirms the regression results, showing a clear upward shift in risk for the stroke-positive group. Similarly, the dot plot of a patient's age versus dementia risk score showed a clear upward trend of increasing dementia risk score as one's age increased (refer to Appendix Figure VI).

## Discussion

In conclusion, the results of this analysis provides supporting evidence for the "Vascular Hypothesis" initially discussed about one's cardiovascular health being a key factor one's likely to get dementia. In particular, this study identified age and history as the key factors associated with higher predicted 1-year dementia risk in the simulated cohort of participants. This makes sense since dementia is age-related condition. However, the baseline demographic model (Model 1) having the weakest predictive power of the models reveals that cogntive decline and dementia is not solely because of aging, but rather influenced by other conditions like one's history of stroke. Thus, this is indicative of cardiovascular intervention needing to be a primary preventative measure against dementia risk in aging patients.

A limitation to keep in mind is the fact that this analysis was done on simulated data with a small sample size of N = 100. It may lack the real-world complexities and randomness that cannot always be simulated. Moreover, the small sample size also means that smaller effects from factors like diabetes may have not been detected easily. The outcome is also a "predicted risk" from prior models and not observed dementia prevalence in patients. So, the results and analysis should not be interpreted casually and generalizability to real-world populations is limited.

In the future, it would be interesting to analyze a larger dataset and analyze how the dementia risk scores change across a decade (since this dataset only captures a 1-year snapshot of predicted risk). This approach would provide more realistic data that doctors and the general population can learn from and adapt to over time for early intervention.

\newpage

## Appendix

### I. Initial Data Cleaning - Missing Value Check

```{r}
#Check for missing values in raw data
na_counts <- colSums(is.na(df_clean))
na_counts[na_counts > 0]
```

### II. Linear Regression Assumption Check for Model 5

```{r}
plot(m5, which = 1)
plot(m5, which = 2)
```

### III. Summary Statistics of Dataset

```{r}
#Get summary statistics of all variables
num <- df_an %>%
  select(age_numeric, dementia_risk) %>%
  summary()

cat <- df_an %>%
  select(sex, hypertension, hyperlipidemia, any_stroke, smoking, t2_diabetes) %>%
  summary()

cat("**Numeric Variables**\n")
kable(num, digits = 4, caption = "Numeric Variable Summary")

cat("Categorical Variables**\n")
kable(cat, caption = "Categorical Variable Summary")
```

### IV. Linear Regression Assumption Check for Model 2

```{r}
plot(m2, which = 1)
plot(m2, which = 2)
```

### V. Boxplot of Stroke History Association with Dementia Risk Score

```{r}
ggplot(df_an, aes(x = any_stroke, y = dementia_risk)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_log10() +
  labs(title = "Stroke History Association with Dementia Risk",
       x = "Presence of Stroke History",
       y = "Dementia Risk Score (log)")
```

### VI. Patient's Age Association with Dementia Risk Score

```{r}
ggplot(df_an, aes(x = age_numeric, y = dementia_risk)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Correlation between Age and Dementia Risk",
       x = "Age (Midpoint of 5-year bracket)",
       y = "Dementia Risk Score")
```

### VII. Model Comparison

```{r}
cat("Model 1")
summary(m1)
cat("Model 2")
summary(m2)
cat("Model 3")
summary(m3)
cat("Model 4")
summary(m4)
cat("Model 5")
summary(m5)
```
